---
title: "Gene Set Enrichment and Clustering"
author: "Fergal Duffy"
date: "2023-11-01"
output: 
  html_document:
    fig_width: 12
    fig_height: 12
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library("tidyverse")
library("ggplot2")
library("ggbeeswarm")
library("ggExtra")
library("fgsea")
library("tmod")
library("msigdbr")
library("DESeq2")
library("pheatmap")
library("igraph")
library("WGCNA")
library("conflicted")
conflicts_prefer(dplyr::filter, base::intersect)
```

## Preparing the data

We will use the Darrah at al dataset to look for gene sets and gene
clusters associated with protection. This experiment comprises 15 rhesus
macaques vaccinated with BCG via various strategies and and BALs
collected at week 13 and 25 post vaccination, prior to Mtb challenge at
week 26.

The Darrah et al dataset is a single-cell RNAseq dataset, and each cell
has been labelled with its canonical named cell type.

To simplify the analysis, cells representing the same cell type have
been summed into a single 'pseudobulk' count per-gene per-sample. Thus,
single-cell RNAseq was used to identify and define cell types, and we
can apply standard bulk RNAseq tools to interpret gene expression within
each cell type.

First, load in the raw count data and associated experiment design
metadata, normalize the counts, tidy up the metadata, and create a long
annotated raw count dataframe.

Key metadata columns

-   VaccineRoute
    -   AE (aerosol BCG)
    -   IDHigh (Intradermal BCG high dose)
    -   IDLow (Intradermal BCG low dose)
    -   IV (intravenous BCG) \*
    -   Naive (unvaccinated control)
-   vaccination\_\_time_since:
    -   N. weeks post vaccine
-   donor_id:
    -   Unique ID per NHP
-   Stimulated
    -   Samples stimulated with PPD: Yes or No

The DESeq2 `vst()` function will normalize the count data to account for
sequencing depth, and the mean-variance relationship, and returns a
log-scaled matrix of normalized counts

```{r}

dataFilename =  '../darrah_etal/pseudo_bulk_wk13_wk25.csv'
metaFilename =  '../darrah_etal/updated_alexandria_metadata.txt'

#Skip the second row, but take column names from the first row
metaCols <- c('organ__ontology_label', 'vaccination', 'vaccination__ontology_label',
              'sex', 'biosample_id', 'cell_type__ontology_label', 'disease__ontology_label', 
              'donor_id', 'Stimulated', 'VaccineRoute',  'VaccineRouteUnique', 
              'VaccineRouteUniqueGroup', 'vaccination_route', 'vaccination__time_since',
              'vaccination__time_since__unit','vaccination__dosage')

metaData <- set_names(read.table(metaFilename, sep="\t", header=F, skip=2), 
                 as.character(read.table(metaFilename,sep="\t",  nrows=1, header=F))) %>%
  select(one_of(metaCols)) %>%
  mutate(countColID=sprintf("D%s_WK%d_STIM%s_%s", donor_id, 
                            vaccination__time_since, Stimulated, make.names(cell_type__ontology_label))) %>%
  distinct()



rawCounts <-  read.csv(dataFilename) %>% 
  rename_with(.fn=\(.)"gene", .cols=1) %>%
  column_to_rownames("gene") %>%
  as.matrix()

longCounts <- rawCounts %>%
  vst() %>%
  data.frame(check.names = F) %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to="countColID", values_to="normExpr") %>%
  left_join(metaData, by="countColID") %>%
  mutate(status=sprintf("%s_%dwks_Stim%s", VaccineRoute, vaccination__time_since, Stimulated))


```

## Differential expression

We can use the standard DESeq2 pipeline to identify gene that are highly
expressed in specific cell types and differ between conditions

```{r}

#Exclude unnamed genes, e.g. LOCxxxx, genexxxx or CXHXorfXX for interpretability
annotatedGenes <- grep("^(gene|LOC|C\\d+H\\d+orf\\d+)", rownames(rawCounts), value=T, invert=T)

#We can fit models incorporating all the study design terms, and extract comparisons from it
#A factor we don't account for: repeated measurements from the same donor_id
studyDesignFormula <- ~0+VaccineRoute+Stimulated+vaccination__time_since


darrahCellTypeDatasets <- unique(metaData$cell_type__ontology_label) %>%
  set_names() %>%
  map(\(cellType){
    message(cellType)
    cellTypeMetadata <- filter(metaData, cell_type__ontology_label==cellType) %>%
      mutate(status=sprintf("%s_%dwks_Stim%s", VaccineRoute, vaccination__time_since, Stimulated),
             countColID=make.names(countColID)) 
    message(cellTypeMetadata$countColID)
    DESeqDataSetFromMatrix(countData=rawCounts[annotatedGenes,cellTypeMetadata$countColID],
                           colData=cellTypeMetadata,
                           design=~0+status)
  })


darrahCelltypeModels <- map(darrahCellTypeDatasets, DESeq)

tCellResponseGenes <-  results(darrahCelltypeModels$`T cell`, 
                               c("status", "IV_13wks_StimYES", "Naive_13wks_StimYES")) %>% 
  data.frame() %>% 
  rownames_to_column("gene") %>% 
  filter(!is.na(padj)) %>% 
  arrange(pvalue)



```

## Exploring Heatmaps and hierarchical clustering

Lets visualize T cell response genes identified above, and how they
respond in other cell types at other timepoints.

We use pheatmap to plot average gene expression per group, and
hierarchically cluster the genes and conditions, indicated with the
dendrogram.

We can also generate clusters manually using the `hclust` function

```{r}

#Differentially expressed genes between IV and naive in stimulated T cells at 13 weeks
sigTcellResponseGenes <- tCellResponseGenes %>%
  filter(abs(log2FoldChange) > 1 & padj < 0.05) %>%
  slice_max(order_by=abs(log2FoldChange), n=50)



#Plot the average expression of response genes per-group in T cells
plotData <- longCounts %>%
  filter(gene %in% sigTcellResponseGenes$gene & cell_type__ontology_label %in% "T cell") %>%
  group_by(gene, status) %>%
  summarize(aveExpr=mean(normExpr, trim=0.25)) %>%
  pivot_wider(names_from=status, values_from=aveExpr) %>%
  column_to_rownames("gene")

columnAnnots <- longCounts %>% 
  select(status, VaccineRoute, vaccination__time_since, Stimulated) %>%
  distinct() %>%
  column_to_rownames("status")

pheatmap(plotData, 
         scale="row",
         annotation_col = columnAnnots)


#Hierarchical clustering done manually
geneClusters <- hclust(dist(plotData), method="ward.D2")
statusClusters <- hclust(dist(t(plotData)), method="ward.D2")

#Pass the clustering to pheatmap
pheatmap(plotData, 
         scale="row",
         annotation_col = columnAnnots,
         cluster_rows = geneClusters,
         cluster_cols = statusClusters)


```

## Gene set enrichment analysis

We can interpret our list of differentially expressed genes by comparing
to annotated empirically derived sets of coherently expressed genes.
Caveat: these gene sets were derived from human whole-blood gene
expression studies, and we are looking at NHP BAL pseudobulk data.
However, human and NHP gene responses are similar, and whole blood
primarily reflects immune cell responses, similar to BAL, so they should
provide useful information.

The `fgsea` packages assesses gene set enrichment using ranked gene
statistics (from our DESeq2 results) and lists of pre-defined gene sets.
We will use MSigDB Hallmark modules, along with published gene sets from
Li et al (LI), and Chaussabel et al (DC).

```{r}
#Format the gene sets
data("tmod") 

liAndDcGsets <- map(tmod$gs2gv, \(gIdx)tmod$gv[gIdx]) %>%
  set_names(paste(tmod$gs$ID, tmod$gs$Title))

hallmarkGsets <- msigdbr(species = "Homo sapiens", category = "H") %>%
  select(gs_name, gene_symbol) %>%
  split(.$gs_name) %>%
  map(\(gsDF)gsDF$gene_symbol)

gsets <- c(liAndDcGsets, hallmarkGsets)

#Filter to only contain genes in our dataset, and drop genesets with < 5 genes
gsets <- gsets %>%
  map(\(gset){
    genes <- intersect(annotatedGenes, gset)
    if (length(genes)>=5) genes else c()
}) %>% compact()

tCellResponseEnrichment <- fgsea(pathways=gsets,
                                 stats=set_names(tCellResponseGenes$log2FoldChange,
                                                 tCellResponseGenes$gene))

sigGeneSets <- tCellResponseEnrichment %>%
  data.frame() %>%
  filter(padj < 0.01) %>%
  arrange(NES) %>%
  mutate(pathway=factor(pathway, levels=pathway))

#Plot the most strongly increased pathway
topPathway <- as.character(slice_max(sigGeneSets, n=1, order_by=NES)$pathway)

plotEnrichment(gsets[[topPathway]],
               set_names(tCellResponseGenes$log2FoldChange,
                                                 tCellResponseGenes$gene)) +
  labs(title=topPathway)

ggplot(sigGeneSets, aes(y=pathway, x=NES, fill=NES)) +
  geom_col() +
  scale_fill_distiller(palette="RdBu") +
  theme_minimal() +
  labs(title="GSEA: BCG-IV vs Naive T cells",
       subtitle="13 weeks, PPD stim", y=NULL)

```

## Unsupervised clustering using WGCNA

WGCNA refers to Weighted Gene Co-expression Network Analysis, created by
Langfelder and Horvath (2018). WGNCA is an unsupervised clustering
approach to find data-driven correlated gene sets in expression data.

The first step in WGCNA is to create an [n_gene x n_gene] correlation
matrix, which is transformed into a adjacency matrix using a
thresholding procedure (hard or soft threshold) applied to the
correlation matrix.

This is followed by an unsupervised gene module detection step. This is
based on topological overlap, e.g. 2 gene have high topological overlap
if they are 'adjacent' to similar sets of genes, based on the adjacency
matrix. This topological overlap measure is used to crease a
hierarchical clustering tree. The clustering tree is then 'cut' into
individual gene modules that are highly topologically similar.

Below, we apply the WGCNA algorithm to the Darrah et al macrophage
expression data to discover and test macrophage-specific gene modules.

#### Identify WGCNA clusters 

```{r}

macNormCountMat <- longCounts %>%
  filter(gene %in% annotatedGenes & cell_type__ontology_label=="macrophage") %>%
  select(gene, countColID, normExpr) %>%
  pivot_wider(names_from="gene", values_from="normExpr") %>%
  column_to_rownames("countColID") %>%
  as.matrix()

powers <-  1:20

softThreshPower <- pickSoftThreshold(
  macNormCountMat,   
  powerVector = powers,
  verbose = 0)

conflicts_prefer(WGCNA::cor)
macWgncaNetwork <- blockwiseModules(datExpr=macNormCountMat, 
                                 power=softThreshPower$powerEstimate,
                                 saveTOMs=T)

macWcgnaLabels <- data.frame(colorLabel=macWgncaNetwork$colors,
                          gene=names(macWgncaNetwork$colors)) %>%
  split(.$colorLabel) %>%
  map(\(df)df$gene)

macWgcnaModuleCounts <- table(macWgncaNetwork$colors)
#One module has the  majority of genes. Since most genes don't respond
#presumably it's the module of non-reponsive genes

#How do the wgcna modules compare with coherently expressed modules
#Lets look at gene level overlaps and test with hypergeometric test
```

#### WGCNA clusters vs published gene sets

We can see how our WGCNA macrophage clusters map to published sets of coherently expressed genes by comparing the WGCNA clusters to published BTM sets.

The heatmap shows signficant overlaps.

```{r}
modulePairs <- expand.grid(WGCNA=names(macWgcnaModuleCounts),
                           BTMs=names(gsets))


macwWcgnaBtmStats <- map2_dfr(modulePairs$WGCNA, modulePairs$BTMs, \(wgcnaName, btmName){
  wgcnaGenes <- macWcgnaLabels[[wgcnaName]]
  btmGenes <- gsets[[btmName]]
  sharedGenes <- intersect(wgcnaGenes, btmGenes)
  pOverlap <- if(length(sharedGenes)==0) {
    1
  } else{
    1 - phyper(length(sharedGenes),
               length(btmGenes),
               length(unique(unlist(gsets))),
               length(wgcnaGenes),
               lower.tail=T)
  }
  data.frame(wgcnaName=wgcnaName, btmName=btmName,
             nWgcnaGenes=length(wgcnaGenes), nBtmGenes=length(btmGenes), nSharedGenes=length(sharedGenes),
             pOverlap=pOverlap)
}) %>% mutate(FDR=p.adjust(pOverlap, method="fdr"))


sigwWcgnaBtmStatsWide <- macwWcgnaBtmStats %>%
  filter(FDR<0.05 & nSharedGenes>=10) %>%
  select(wgcnaName, btmName, nSharedGenes) %>%
  pivot_wider(names_from="btmName", values_from=nSharedGenes, values_fill = 0) %>%
  column_to_rownames("wgcnaName")

pheatmap(sigwWcgnaBtmStatsWide, 
         color = RColorBrewer::brewer.pal(n=7, name="YlOrRd"),
         clustering_method="ward.D2")

```

#### How do WGCNA clusters relate to experimental status?

Averaging WGCNA module expression by status lets us visualize relationships
between WGNCA macrophage clusters and mutliple aspects of the experiments immune
responses

```{r}
#Now, average WGCNA module by cell type and vaccination status

WGNAmodStats <- longCounts %>%
  filter(cell_type__ontology_label %in% "macrophage") %>%
  left_join(data.frame(gene=names(macWgncaNetwork$colors),
                       WGCNAmod=macWgncaNetwork$colors)) %>%
    filter(!is.na(WGCNAmod)) %>%
  group_by(WGCNAmod) %>%
  summarize(aveModExpr=mean(normExpr, trim=0.25), sdModExpr=sd(normExpr))
  

aveWGNAexprByStatusAndCellType <- longCounts %>%
    filter(cell_type__ontology_label %in% "macrophage")%>%
  left_join(data.frame(gene=names(macWgncaNetwork$colors),
                       WGCNAmod=macWgncaNetwork$colors)) %>%
  filter(!is.na(WGCNAmod)) %>%
  group_by(WGCNAmod, status, cell_type__ontology_label) %>%
  summarize(modExprPerStatusPerCellType=mean(normExpr, trim=0.25)) %>%
  ungroup() %>%
  left_join(WGNAmodStats, by="WGCNAmod") %>%
  mutate(scaledModExprPerStatusPerCellType = (modExprPerStatusPerCellType-aveModExpr)/sdModExpr)

statusAnnots <- longCounts %>%
  select(celltype=cell_type__ontology_label, status, VaccineRoute,
         timepoint=vaccination__time_since, stim=Stimulated) %>%
  distinct() %>%
  mutate(colName=paste(celltype, status, sep="|"), timepoint=factor(timepoint)) %>%
  select(-status, -celltype) %>%
  column_to_rownames("colName")

aveWGNAexprByStatusAndCellTypeWide <- aveWGNAexprByStatusAndCellType %>%
  select(WGCNAmod, status, cell_type__ontology_label, scaledModExprPerStatusPerCellType) %>%
  mutate(label=paste(cell_type__ontology_label, status, sep="|")) %>%
  select(label,WGCNAmod, scaledModExprPerStatusPerCellType) %>%
  pivot_wider(names_from=label, values_from=scaledModExprPerStatusPerCellType) %>%
  column_to_rownames("WGCNAmod")

annotColors <- list(celltype=c("B cell"="purple",
                              "eosinophil"="yellow",
                              "epithelial cell"="grey",
                              "macrophage"="green",
                              "mast cell"="orange",
                              "neutrophil"="red",
                              "T cell"="black"),
                    stim=c(NO="white", YES="black"),
                    timepoint=c(`13`="white", `25`="grey"),
                    VaccineRoute=c(Naive="white", AE="purple", 
                                   IDHigh="darkgreen", IDLow="green", 
                                   IV="firebrick"))

pheatmap(aveWGNAexprByStatusAndCellTypeWide,
         annotation_col = statusAnnots,
         show_colnames = F, annotation_colors = annotColors,
         clustering_method = "ward.D2")

```
